Overlay databank unlocks data-driven analyses of biomolecules for all

### 2023-10-11 SI.tex
Update text

\pagebreak
\section{Testing machine learning models against literature data}
\begin{figure}[hb]
    \centering
    \includegraphics[width=50mm]{Figures/MLtest/Ridgetest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/Lassotest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/ElasticNettest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/DecisionTreeRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/RandomForestRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/GradientBoostingRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/AdaBoostRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/KNeighborsRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/MLPRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/XGBRegressortest.pdf}
    \caption{Predictions of area per lipids of POPC, POPE, POPS, PI, sphingomyelin lipids and cholesterol from linear regression model against literature data from simulations (green~\cite{shahane19}, blue~\cite{kumar21}, and red~\cite{oliveira22}).}
    \label{fig:LinearRegressionTest}
\end{figure}

>>>>

\pagebreak
\section{Testing machine learning models against literature data}
\begin{figure}[hb]
    \centering
    \includegraphics[width=50mm]{Figures/MLtest/Ridgetest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/Lassotest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/ElasticNettest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/DecisionTreeRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/RandomForestRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/GradientBoostingRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/AdaBoostRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/KNeighborsRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/MLPRegressortest.pdf}
    \includegraphics[width=50mm]{Figures/MLtest/XGBRegressortest.pdf}
    \caption{Predictions of area per lipids of POPC, POPE, POPS, PI, sphingomyelin lipids and cholesterol from the tested machine learning models against literature data from simulations (green~\cite{shahane19}, blue~\cite{kumar21}, and red~\cite{oliveira22}).}
    \label{fig:MLTest}
\end{figure}

### 2023-10-11 main.tex
Updates from Overleaf
Update text

 After the parameter optimization, we tested predictions both against the remaining 20\% of the databank and against data from different ML models against area per lipids reported from simulations of membranes containing mixtures of POPC, POPE, POPS, PI, sphingomyelin lipids and cholesterol~\cite{shahane19,kumar21,oliveira22} not included in the databank. Linear regression and Ridge models gave the best correlation with the literature data in this test (predictions  from the linear regression model are shown in Fig.~\ref{fig:LinearRegressionTest}). The linear regression model was selected for further studies due to its simplicity.
\begin{figure}
    \centering
    \includegraphics[width=88mm]{Figures/LinearRegressionTest.pdf}
    \caption{Predictions of area per lipids of POPC, POPE, POPS, PI, sphingomyelin lipids and cholesterol from linear regression model against literature data from simulations (green~\cite{shahane19}, blue~\cite{kumar21}, and red~\cite{oliveira22}).}
    \label{fig:LinearRegressionTest}
\end{figure}

To demonstrate the usefulness of the constructed models for understanding multi-component membrane properties, we predicted later packing (ares per lipids) and membrane thicknesses of common biological membranes based on their lipid compositions reported in the literature, see table~\ref{tab:MLpredictions}. The model predicts substantial 50\% difference in area per lipid and 0.5\,nm in thickness between most densely (influenza virus) and loosely (mitochondria) packed membranes. Because these parameters correlate with many biologically relevant functions of membranes, such as permeation, cholesterol flip-flops (see next sections), and interactions with proteins~\cite{phillips09}, our results demonstrate that NMRlipids databank can be used to give valuable insights on biologically relevant properties of complex biological membranes. Most importantly, the delivered programmatic access to increasing amount of MD simulation data enables training of machine learning models that predict various membrane properties for all.

    \begin{table}[htb]
    \caption{Area per lipids (APL) and membrane thicknesses predicted by the linear regression model trained using the NMRlipids databank for membrane compositions corresponding different biological membranes. Systems are sorted in the order of increasing membrane packing. Compositions of different membranes are estimated based on references given in composition column.}
        \centering
        \begin{tabular}{cccc}
             Membrane&  Composition [lipid(\%-fraction)]&  APL[\AA$^2$] & Thickness [nm$^2$]\\
             \hline
             Mitochondria & POPC(37) : POPE(31) : PI(6) : CL(22) : 
             CHOL(4)~\cite{vanmeer08,escriba15,casares19} & 74 (66)$^*$ & 4.3 \\
             Bacterial & POPC(20) : POPE(35) : POPG(35) : CL(5) : CHOL(5)~\cite{chwastek20}  & 62 & 4.2 \\
             ER &  POPC(54) : POPE(20) : POPS(4) : PI(11) : SM(4) : CHOL(8)~\cite{vanmeer08,escriba15,casares19} & 56 & 4.5 \\
             Golgi & POPC(36) : POPE(21) : POPS(6) : PI(12) : SM(7) : CHOL(18)~\cite{vanmeer08,escriba15,casares19} & 51 & 4.7 \\
             Plasma & POPC(23) : POPE(11) : POPS(8) : PI(7) : SM(17) : CHOL(34)~\cite{vanmeer08,escriba15,casares19} & 45 & 4.9 \\
             Synaptic & POPC(28) : POPE(20) : POPS(9) : PI(2) : SM(4) : CHOL(37)~\cite{binotti21}  &  45 & 4.8 \\
             Influenza & POPC(5) : POPE(32) : POPS(15) : SM(9) : CHOL(40)~\cite{gerl12,ivanova15} & 42 & 4.8 \\
        \end{tabular}
        \label{tab:MLpredictions}

@@

\begin{table}[t]
    \centering
    \begin{tabular}{p{3.0cm}  p{4.0cm}  p{7.0cm}}
     Target group  &  Outcomes     & Practical examples \\
    \hline
    Computational and data scientists & Databanks with programmatic access, training sets for machine learning applications, automatic optimization of simulation models & Programmatic access to all publicly available MD simulation trajectories, machine learning models predicting properties of complex biomolecular assemblies, optimizing force field and model parameters from atomistic to continuum scale \\

@@

\subsection{Universal naming convention for molecules and atoms} \label{naming}
When analysing simulation trajectories, atoms and molecules often need to be  called by the names used in the trajectory. However, these names typically vary between force fields, as a universal naming convention has not been defined for lipids. To enable automatic analyses over all the simulations in the NMRlipids Databank, we have defined universal naming conventions for the molecules and atoms therein. The universal abbreviations used in the NMRlipids Databank for each molecule are listed in the NMRlipids databank documentation \href{https://nmrlipids.github.io/moleculesAndMapping.html}{nmrlipids.github.io/moleculesAndMapping.html}.
%Table~\ref{tab:abbreviations} in the supplementary information. 
The atom names used in simulation trajectories are connected to the universal atom names using the mapping files defined in the NMRlipids Project (\url{https://nmrlipids.github.io/moleculesAndMapping.html}). These files are located at \texttt{/Scripts/BuildDatabank/mapping\_files} in the NMRlipids Databank repository. These files also define whether an atom belongs to the headgroup, glycerol backbone, or acyl chain region in a lipid. In practise, the force-field-specific molecule names and mapping file
names are given in the COMPOSITION dictionary in the README.yaml files for each molecule in each simulation in the NMRlipids Databank.

\subsection{Adding data to NMRlipids Databank}
The NMRlipids Databank is open for additions of simulation data by anyone. In practise, the required information is first manually entered into an info.yaml file that is then added into the \texttt{/Scripts/BuildDatabank/info\_files} folder trough a git pull request. Rest of the information to be stored in the README.yaml files will be then automatically extracted using the \texttt{/Scripts/BuildDatabank/AddData.py} script. The required manually entered and automatically extracted information are described in detail in Table~\ref{tab:READMEkeys} in the supplementary information and in the NMRlipids databank documentation \href{https://nmrlipids.github.io/}{nmrlipids.github.io/}. The documentation includes also detailed instructions to add data. To avoid ineligible entries and minimize human errors, the pull requests are monitored before the acceptance and generation of the README.yaml files.  Currently, the NMRlipids Databank is composed of simulations that are found from the Zenodo repository with an appropriate license; most, but not all, of these trajectories originate from previous NMRlipids projects~\cite{botan15,catte16,antila19,bacle21}.

\subsection{Experimental data}
Experimental data used in the quality evaluation, currently composed of C--H bond order parameters and X-ray scattering form factors, are stored in \texttt{/Data/experiments} in the NMRlipids Databank repository. Similarly to simulations, each experimental data set has a README.yaml file containing all the relevant information about the experiment. The keys and their descriptions for the experimental data, as well as detailed instructions to add the date, are given in Table~\ref{tab:READMEkeysEXP} and in the NMRlipids databank documentation \href{https://nmrlipids.github.io/}{nmrlipids.github.io}. The NMR data currently in the NMRlipids Databank are taken from Refs.~\citenum{scherer87,ferreira13,antila19,melcr20,bacle21,hanashima19,doktorova20} and the X-ray scattering data from Refs.~\citenum{Kucerka08a,kucerka11,pan12b,pan14,kucerka15,NMRlipidsIII,doktorova20}. In addition, previously unpublished NMR data for POPE, POPG, and DOPC was acquired as described in the supplementary information (Figs.~\ref{POPEexp1}--\ref{fig:DOPCexp}) and contributed to the Databank.

\subsection{Analysing simulations}
In practise, simulations in the NMRlipids Databank can be analyzed by executing a program that
1) loops over the README.yaml files in the {\it Databank layer},
2) downloads the data using the information in the README.yaml files, and then
3) performs the desired analysis on a local computer utilising the universal naming conventions for molecules and atoms defined in the README.yaml and mapping files.
This general procedure is illustrated in Fig.~\ref{fig:overlay}D, and a template for user-defined analyses is available at at \href{https://github.com/NMRLipids/databank-template}{github.com/NMRLipids/databank-template}. Further practical examples of codes  performing such analyses are listed in Tables~\ref{tab:codes} and~\ref{tab:codesApplication}.  The equilibration period given by the user (TIMELEFTOUT in Table~\ref{tab:READMEkeys}) is discarded from the trajectories in all analysis codes. For further details, see the NMRlipids databank documentation \href{https://nmrlipids.github.io/}{nmrlipids.github.io}.

@@

\subsection{Calculation of area per lipid and bilayer thickness}
Area per lipids of bilayers were calculated by dividing the time-averaged area of the simulation box with the total number of lipids and surfactant molecules in the simulation (see Table~\ref{tab:abbreviations} for the list of molecules considered as lipids or surfactants). The script that calculates the area per lipid from all simulations in the NMRlipids Databank repository is available at \texttt{Scripts/AnalyzeDatabank/calcAPL.py} in the NMRlipids Databank repository. The resulting area per lipids are stored for all simulations in files named \texttt{apl.json} at folders in \texttt{/Data/Simulations}.

@@

\subsection{Training machine learning model to predict properties of multi-component membranes}
    Different ML approaches were tested to find the best model to predict the area per lipids of multi-component membranes  from their lipid composition. To his end, we tested linear, Lasso, Ridge, elastic net, decision tree, random forest, multi-layer perceptron, k-nearest neighbour, gradient boost, AdaBoost, and XGBoost regressors. The models were trained to predict the area per lipid for a membrane with the given molar fractions of membrane lipids. Most abundant lipids in the databank, POPC, POPE, POPG, POPS and cholesterol, were considered explicitly, while SM, cardiolipins and PI lipids with  more sparse data were all grouped together based on their headgroups independently on acyl chain content. Salt concentration, temperature and other conditions were not considered as parameters at this stage.

To find the optimal model for this task, we first randomly divided simulations into training (80\% of the simulations) and test (20\% of the simulations) sets. Then, we used the training set with 5-fold cross validation to conduct a grid search for hyperparameters for a set of ML regression models. The best performing hyperparameters were then used to fit each model to the training set. The performance of the models was then tested on the remaining 20\% of the data extracted from the databank, and additionally, 
%selected optimal parameters for each model based on their performance in predicting area per lipids for simulations in the test set. 
%After finding optimal parameters for each ML model, each model was trained again using all simulations in the databank. 
%The performances of newly trained models were then tested 
by predicting area per lipids from multi-component membrane simulations reported in the literature but not included in the databank~\cite{shahane19,kumar21,oliveira22}. Due to its  simplicity and good performance in these tests, we selected to use linear regression model for further studies. The  linear regression model predicting membrane thicknesses was trained in similarly. The ML models and analysis were implemented with Python using scikit-learn which was supplemented with the xgboost library to include extreme gradient boosted decision tree regressor to the set of tested model.

>>>>

 After the parameter optimization, we tested predictions both against the remaining 20\% of the databank and against data from different ML models against area per lipids reported from simulations of membranes containing mixtures of POPC, POPE, POPS, PI, sphingomyelin lipids and cholesterol~\cite{shahane19,kumar21,oliveira22} not included in the databank. Linear regression and Ridge models gave the best correlation with the literature data in this test (predictions  from the linear regression model are shown in Fig.~\ref{fig:LinearRegressionTest} and from other models in Fig.~\ref{fig:MLTest} in the supplementary information). The linear regression model was selected for further studies due to its simplicity.
\begin{figure}
    \centering
    \includegraphics[width=88mm]{Figures/MLtest/LinearRegressiontest.pdf}
    \caption{Predictions of area per lipids of POPC, POPE, POPS, PI, sphingomyelin lipids and cholesterol from linear regression model against literature data from simulations (green~\cite{shahane19}, blue~\cite{kumar21}, and red~\cite{oliveira22}).}
    \label{fig:LinearRegressionTest}
\end{figure}

To demonstrate the usefulness of the constructed models for understanding multi-component membrane properties, we predicted later packing (ares per lipids) and membrane thicknesses of common biological membranes based on their lipid compositions reported in the literature, see table~\ref{tab:MLpredictions}. The model predicts substantial 50\% difference in area per lipid between most densely (influenza virus) and loosely (mitochondria) packed membranes. Difference of 0.8\,nm in thickness is predicted between thinnest (bacterial) and thicknest (plasma) membranes. Such differences are expected to effect on many biologically relevant functions of membranes, such as permeation, cholesterol flip-flops (see next sections), and interactions with proteins~\cite{phillips09}, demonstrating that NMRlipids databank can be used to give valuable insights on biologically relevant properties of complex biological membranes. Most importantly, the delivered programmatic access to increasing amount of MD simulation data enables training of machine learning models that predict various membrane properties for all.

    \begin{table}[htb]
    \caption{Area per lipids (APL) and membrane thicknesses predicted by the linear regression model trained using the NMRlipids databank for membrane compositions corresponding different biological membranes. Systems are sorted in the order of increasing membrane packing. Compositions of different membranes are estimated based on references given in composition column.}
        \centering
        \begin{tabular}{cccc}
             Membrane&  Composition [lipid(\%-fraction)]&  APL[\AA$^2$] & Thickness [nm$^2$]\\
             \hline
             Mitochondria & POPC(37) : POPE(31) : PI(6) : CL(22) : 
             CHOL(4)~\cite{vanmeer08,escriba15,casares19} & 74 (66)$^*$ & 4.4 \\
             Bacterial & POPC(20) : POPE(35) : POPG(35) : CL(5) : CHOL(5)~\cite{chwastek20}  & 61 (60)$^*$ & 4.2 \\
             ER &  POPC(54) : POPE(20) : POPS(4) : PI(11) : SM(4) : CHOL(8)~\cite{vanmeer08,escriba15,casares19} & 57 & 4.6 \\
             Golgi & POPC(36) : POPE(21) : POPS(6) : PI(12) : SM(7) : CHOL(18)~\cite{vanmeer08,escriba15,casares19} & 52 & 4.8 \\
             Plasma & POPC(23) : POPE(11) : POPS(8) : PI(7) : SM(17) : CHOL(34)~\cite{vanmeer08,escriba15,casares19} & 45 & 5.0 \\
             Synaptic & POPC(28) : POPE(20) : POPS(9) : PI(2) : SM(4) : CHOL(37)~\cite{binotti21}  &  45 & 4.8 \\
             Influenza & POPC(5) : POPE(32) : POPS(15) : SM(9) : CHOL(40)~\cite{gerl12,ivanova15} & 42 & 4.8 \\
        \end{tabular}
        \label{tab:MLpredictions}

@@

\begin{table}[t]
    \centering
    \begin{tabular}{p{3.0cm}  p{4.0cm}  p{7.0cm}}
     Target group  &  Outcomes     & Practical examples \\
    \hline
    Computational and data scientists & Databanks with programmatic access, training sets for machine learning applications, automatic optimization of simulation models & Programmatic access to all publicly available MD simulation trajectories, machine learning models predicting properties of complex biomolecular assemblies (Table~\ref{tab:MLpredictions}), optimizing force field and model parameters from atomistic to continuum scale \\

@@

\subsection{Universal naming convention for molecules and atoms} \label{naming}
When analysing simulation trajectories, atoms and molecules often need to be  called by the names used in the trajectory. However, these names typically vary between force fields, as a universal naming convention has not been defined for lipids. To enable automatic analyses over all the simulations in the NMRlipids Databank, we have defined universal naming conventions for the molecules and atoms therein. The universal abbreviations used in the NMRlipids Databank for each molecule are listed in the NMRlipids Databank documentation~\cite{NMRlipidsDocsMolAndMapping}.
%Table~\ref{tab:abbreviations} in the supplementary information. 
The atom names used in simulation trajectories are connected to the universal atom names using the mapping files defined in the NMRlipids Project~\cite{NMRlipidsDocsMolAndMapping}. These files are located at \texttt{/Scripts/BuildDatabank/mapping\_files} in the NMRlipids Databank repository. These files also define whether an atom belongs to the headgroup, glycerol backbone, or acyl chain region in a lipid. In practise, the force-field-specific molecule names and mapping file
names are defined in the README.yaml files for each molecule in each simulation as described in the NMRlipids Databank documentation~\cite{NMRlipidsREADMEfiles}.

\subsection{Adding data to NMRlipids Databank}
The NMRlipids Databank is open for additions of simulation data by anyone. In practise, the required information is first manually entered into an info.yaml file that is then added into the \texttt{/Scripts/BuildDatabank/info\_files} folder trough a git pull request. Rest of the information to be stored in the README.yaml files will be then automatically extracted using the \texttt{/Scripts/BuildDatabank/AddData.py} script. The required manually entered and automatically extracted information are described in detail in Table~\ref{tab:READMEkeys} in the supplementary information and in the NMRlipids databank documentation~\cite{NMRlipidsREADMEfiles}. The documentation includes also detailed instructions to add data. To avoid ineligible entries and minimize human errors, the pull requests are monitored before the acceptance and generation of the README.yaml files.  Currently, the NMRlipids Databank is composed of simulations that are found from the Zenodo repository with an appropriate license; most, but not all, of these trajectories originate from previous NMRlipids projects~\cite{botan15,catte16,antila19,bacle21}.

\subsection{Experimental data}
Experimental data used in the quality evaluation, currently composed of C--H bond order parameters and X-ray scattering form factors, are stored in \texttt{/Data/experiments} in the NMRlipids Databank repository. Similarly to simulations, each experimental data set has a README.yaml file containing all the relevant information about the experiment. The keys and their descriptions for the experimental data, as well as detailed instructions to add the date, are given in Table~\ref{tab:READMEkeysEXP} and in the NMRlipids databank documentation~\cite{NMRlipidsExperimentalData}. The NMR data currently in the NMRlipids Databank are taken from Refs.~\citenum{scherer87,ferreira13,antila19,melcr20,bacle21,hanashima19,doktorova20} and the X-ray scattering data from Refs.~\citenum{Kucerka08a,kucerka11,pan12b,pan14,kucerka15,NMRlipidsIII,doktorova20}. In addition, previously unpublished NMR data for POPE, POPG, and DOPC was acquired as described in the supplementary information (Figs.~\ref{POPEexp1}--\ref{fig:DOPCexp}) and contributed to the Databank.

\subsection{Analysing simulations}
In practise, simulations in the NMRlipids Databank can be analyzed by executing a program that
1) loops over the README.yaml files in the {\it Databank layer},
2) downloads the data using the information in the README.yaml files, and then
3) performs the desired analysis on a local computer utilising the universal naming conventions for molecules and atoms defined in the README.yaml and mapping files.
This general procedure is illustrated in Fig.~\ref{fig:overlay}D, and a templates for user-defined analyses are available via NMRlipids Databank documentation~\cite{NMRlipidsExampleAndTutorials}. Further practical examples of codes performing such analyses are listed in Tables~\ref{tab:codes} and~\ref{tab:codesApplication}.  The equilibration period given by the user (TIMELEFTOUT in Table~\ref{tab:READMEkeys}) is discarded from the trajectories in all analysis codes. For further details, see the NMRlipids databank documentation \href{https://nmrlipids.github.io/}{nmrlipids.github.io}.

@@

\subsection{Calculation of area per lipid and bilayer thickness}
Area per lipids of bilayers were calculated by dividing the time-averaged area of the simulation box with the total number of lipids and surfactant molecules in the simulation. 
%(see Table~\ref{tab:abbreviations} for the list of molecules considered as lipids or surfactants). 
The script that calculates the area per lipid from all simulations in the NMRlipids Databank repository is available at \texttt{Scripts/AnalyzeDatabank/calcAPL.py} in the NMRlipids Databank repository. The resulting area per lipids are stored for all simulations in files named \texttt{apl.json} at folders in \texttt{/Data/Simulations}.

@@

\subsection{Training machine learning model to predict properties of multi-component membranes}
Different machine learning approaches were tested to find the best model to predict the area per lipids of multi-component membranes  from their lipid composition. To his end, we tested linear, Lasso, Ridge, elastic net, decision tree, random forest, multi-layer perceptron, k-nearest neighbour, gradient boost, AdaBoost, and XGBoost regressors. The models were trained to predict the area per lipid for a membrane with the given molar fractions of membrane lipids. Most abundant lipids in the databank, POPC, POPE, POPG, POPS and cholesterol, were considered explicitly, while SM, cardiolipins and PI lipids with  more sparse data were all grouped together based on their headgroups independently on acyl chain content. Salt concentration, temperature and other conditions were not considered as parameters at this stage.

To find the optimal model for this task, we first randomly divided simulations into training (80\% of the simulations) and test (20\% of the simulations) sets. Then, we used the training set with 5-fold cross validation to conduct a grid search for hyperparameters for a set of ML regression models. The best performing hyperparameters were then used to fit each model to the training set. The performance of the models was then tested on the remaining 20\% of the data extracted from the databank, and additionally, by predicting area per lipids from multi-component membrane simulations reported in the literature but not included in the databank~\cite{shahane19,kumar21,oliveira22}. Due to its  simplicity and good performance in these tests, we selected to use linear regression model for further studies. The  linear regression model predicting membrane thicknesses was trained in similarly. The ML models and analysis were implemented with Python using scikit-learn which was supplemented with the xgboost library to include extreme gradient boosted decision tree regressor to the set of tested model. Jyputer notebook that trains the models and predicts are per lipids is available at \texttt{scripts/APLpredictor.ipynb} in the repository at~\url{https://github.com/NMRLipids/DataBankManuscript/}. In further applications, it is important to consider potential limitations arising from the currently limited amount of data for certain types of systems. For example, the current models have been trained and tested only for mixtures with palmitoyl and oleyl acyl chains because the databank does not yet contain enough data with varying number of double bonds. Such limitations are expected to decrease with increasing amount of data in the future.
